---
title: Troubleshooting and asking for help
subtitle: 'With coding comes bugs and errors that need troubleshooting. In this chapter, you will learn how to decipher errors, find help for fixing them, and practice asking clear questions with minimally reproducible examples.'
title-block-banner: yes
engine: knitr
filters:
  - webr-teachr
webr:
  packages: ["tidyverse"]
---

```{r setup, include = FALSE}
library(tidyverse)
knitr::knit_hooks$set(class = function(before, options, envir) {
  if(before) {
    sprintf("<div class='%s'>", options$class)
  } else {
    "</div>"
  }
})
knitr::opts_chunk$set(fig.retina = 2,
                      fig.width = 6, 
                      fig.height = 4,
                      fig.align = "center",
                      cache.path = "cache/",
                      message = FALSE,
                      warning = FALSE)
library(icons)

detach_package <- function(pkg, character.only = FALSE)
{
  if(!character.only)
  {
    pkg <- deparse(substitute(pkg))
  }
  search_item <- paste("package", pkg, sep = ":")
  while(search_item %in% search())
  {
    detach(search_item, unload = TRUE, character.only = TRUE)
  }
}

```

# What are error messages?

Eek! There's a bug `r icon_style(fontawesome("bug", style = "solid"),fill = "red")` in my code and it's causing lots of <span style="color: red;">scary red errors</span>!

Don't panic - believe it or not, this error is just trying to help you! It also may have just saved you from invalid results that you may not have have noticed.

Errors can be frustrating, essentially you've just asked the computer to do something and it has said no!

There are a few troubleshooting techniques that we can use to find the source of the error, modify your code, and get the results you're looking for.

# Read the error

The first step is to carefully **read** the error message, it often contains clues into where and why the code didn't run. With some time and experience, you'll begin to identify these errors and fix your code quickly based on the message alone!

## Let's have a look one of the most common error message 

```{webr-teachr}
# One of the r built-in dataset is called "cars"
# Let's try to have a look at the structure of the cars dataset 
str(car)

```

Oh no! An error message: <span style="color: red;">"Error: object 'car' not found"errors</span>!

When the error message says that <span style="color: red;">"Error: object 'something' not found"errors</span>!, this is often the case that there might be a typo or you didn't load the object loaded in correctly. Let's see how we could fix this. 

```{webr-teachr}
# One of the r built-in dataset is called "cars"
# Let's try to have a look at the structure of the cars dataset 
str(<<cars>>)
  
???
  
p <- .code[[1]][[2]][[2]]
  
c("did you put in cars?" = !"cars" %in% all.vars(.code))
```

Typo is a very common error that you could make. This could be a miss-type of the object name or miss-type of function name. When you see error messages suggesting <span style="color: red;">"Error: object 'something' not found"errors</span>!, please check and make sure that you have correctly typed everything. 


## What if we cannot find the function?!

```{r,echo = FALSE, eval=TRUE}
detach_package(tidyverse)
```

```{webr-teachr}
# One of the r built-in dataset is called "cars"
# Let's try to have a look at the structure of the cars dataset 
# Instead of the base R function of checking structure for your dataset, let's use tidyverse function glance to do that. 

glimpse(cars)
  
???
  
```

Oh no, we could not find the function of `glance()`! We made sure that the spelling is correct, what could be the problem now?

When you see the error messages says that <span style="color: red;">"Error: could not find function "something" "errors</span>!,it normally means that the required package hasn't been loaded to the environment.

Let's try add in the required package `tidyverse`



```{webr-teachr}
# One of the r built-in dataset is called "cars"
# Let's try to have a look at the structure of the cars dataset 
# Instead of the base R function of checking structure for your dataset, let's use tidyverse function glance to do that. 

library(<<tidyverse>>)
glimpse(cars)
  
???
  
```

